---
title: "top 2000 vs top 50"
author: "Laurens de Vries"
date: "2/15/2021"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(spotifyr)

```

The Top 2000 is an annual Dutch marathon radio program, that plays the 2,000 most popular songs of all time determined by a public vote. The highest ranking songs do not vary much from year to year but there sometimes new songs. I am interested in if these new songs are similar to older high ranking songs compared to other popular songs from the same year.

The new top 2000 songs probably have more focus on instruments compared to other hit songs from the same years.

As for the corpus i will be using the 2020 edition of the top 2000 and comparing to the spotify Dutch top 50 playlist.



The following graphs will compare the distribution off certain variables between the playlists.
```{r, include = FALSE}
top2000 <- get_playlist_audio_features("", "1DTzz7Nh2rJBnyFbjsH1Mh")
top50 <- get_playlist_audio_features("", "37i9dQZEVXbKCF6dqVpDkS")

awards <-
  bind_rows(
    top2000 %>% mutate(category = "Top 2000"),
    top50 %>% mutate(category = "Top 50")
  )

```

```{r}
awards %>%
  ggplot(aes(x = category, y = energy)) +
  geom_violin()

awards %>%
  ggplot(aes(x = category, y = acousticness)) +
  geom_violin()

awards %>%
  ggplot(aes(x = category, y = valence)) +
  geom_violin()

```


